Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-01 rnni_convexity.tex
pass section cluster thm

Within this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
Within this section we will explain why this is not the case.
Therefore we consider an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness and explain why this example cannot be transferred to $\rnni$.
Following this, we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.

@@

\begin{theorem}
There are trees $T,R$ in $\nni$ sharing a cluster which is not shared by any intermediate tree on any shortest path from $T$ to $R$.
\label{thm:split_nni}
%The formulation of this lemma is a bit different in Li1996-zw.
%However, the lemma as stated here follows directly from Li1996-zw.
\end{theorem}

@@

In the proof presented in \autocite{Li1996-zw} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
We describe this example in the following a simplified illustration of it can be found in Figure~\ref{fig:NNI_NP_proof}.
In the example of \autocite{Li1996-zw} trees $T$ and $R$ on $2n$ taxa are considered where each of the trees consist of two caterpillar trees that share the root as parent.
The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $n$ cherries, each of which contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is transformed to a tree $R'$ containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, \autocite{Li1996-zw} use the fact that $d_{\nni}(T',R') = d_{\nni}(T{\big|}_n, R{\big|}_n)$ where $T{\big|}_n$ and $R{\big|}_n$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively.
For $\rnni$ there are lots of additional rank moves needed on a path that follows the principle of the shortest path in $\nni$ as described above.
Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, this results in a path that is longer than a shortest $\rnni$ path.
Therefore, a path where the two caterpillar subtrees of $T$ and $R$ are sorted separately is shorter in $\rnni$.

\begin{figure}[H]
\centering
\includegraphics[width=.8\textwidth]{NNI_NP_proof}
\vspace{12pt}
\caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
\label{fig:NNI_NP_proof}
\end{figure}

Since the example given in \autocite{Li1996-zw}, which can be used to prove Theorem~\ref{thm:split_nni} in $\nni$, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
For understanding this conjecture it is important to know the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
By deleting an edge $e$ of a tree $T$ one receives two graphs with labels in $A$ and $B$, respectively, such that $\{A,B\}$ is a partition of the taxon set of $T$.
Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
Since these partitions are commonly referred to as splits, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.

@@

Since the counterexample in Figure~\ref{fig:splitthm_counterexample} shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem} (Conjecture~\ref{conjecture:cluster_theorem}).
Within the Cluster Theorem we are considering clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
Note that trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but share no cluster.

@@

For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.

>>>>

In this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.

@@

\begin{theorem}
There are trees $T,R$ in $\nni$ sharing a cluster which is not shared by any intermediate tree on any shortest path from $T$ to $R$.
\label{thm:split_nni}
\end{theorem}

@@

In the proof of Theorem~\ref{thm:split_nni} presented in \autocite{Li1996-zw} an example of two trees $T$ and $R$ that share clusters $C$ and $X \setminus C$ but have no shortest path that preserves these cluster is given.
A simplified illustration of this example is provided in Figure~\ref{fig:NNI_NP_proof}.
\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $\frac{n}{2}$ cherries, each of which contains one taxon of each of the clusters $C$ and $X \setminus C$.
Then $T'$ is sorted to a tree $R'$ containing the same cherries as $T'$ such that that resolving all these cherries of $R'$ results in $R$.
Specifically, \autocite{Li1996-zw} prove that there is no shortest path between $T$ and $R$ in $\nni$ where $C$ and $X \setminus C$ are present in all trees on that path.

The reason that this example cannot be transferred to $\rnni$ is that there are lots of additional rank moves needed on a path between $T$ ans $R$ similar to the one described above.
As in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is longer than a shortest $\rnni$ path.
Sorting the two caterpillar subtrees of $T$ and $R$ separately results in a shorter path in $\rnni$.

\begin{figure}[H]
\centering
\includegraphics[width=.8\textwidth]{NNI_NP_proof}
\vspace{12pt}
\caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
In \autocite{Li1996-zw} it is proven that there is a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
Instead, the depicted path where $\frac{n}{2}$ cherries are built and then sorted before being resolved is a shortest path}
\label{fig:NNI_NP_proof}
\end{figure}

Since the example of a shortest $\nni$ described above, which was used to prove Theorem~\ref{thm:split_nni} in \autocite{Li1996-zw}, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
For understanding this conjecture it is important to know the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
By deleting an edge $e$ of a tree $T$ one receives two graphs with labels in $A$ and $B$, respectively, such that $\{A,B\}$ is a partition of the taxon set of $T$.
Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
Since these partitions are commonly referred to as splits, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.

@@

Since the counterexample in Figure~\ref{fig:splitthm_counterexample} shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem} (Conjecture~\ref{conjecture:cluster_theorem}).
The Cluster Theorem concerns clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
Note that trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but share no cluster.

@@

For providing further evidence that the Cluster Theorem holds in $\rnni$, we could show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this subgraph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
